---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE, warning=FALSE} 
library(lubridate)
file <- "activity.csv"
data <- read.csv(file, header = TRUE, stringsAsFactor = FALSE)
data$date <- ymd(data$date)
```

## What is the mean number of steps taken per day?
```{r mean_median, echo=TRUE}
steps_day <- aggregate(steps~date, data, sum)    
hist(steps_day$steps, main = "Steps Taken Per Day", col = "turquoise", xlab = "steps/day")
mean(steps_day$steps)  
median(steps_day$steps) 
```   

## What is the average daily activity pattern?
```{r echo=TRUE}
steps_int <- aggregate(steps~interval, data, mean, na.rm=TRUE)
maxActiveInt <- steps_int[which.max(steps_int$steps),1]
maxSteps <-  round(max(steps_int$steps))
```
```{r dailyplot, echo=TRUE}
plot(steps_int$interval, steps_int$steps, type = "l", xlab = "5-minute interval", ylab = "steps")
```  

The 5 minute interval with the maximun number of steps is interval **`r maxActiveInt`** with an average of **`r maxSteps`** of steps taken.  

## Imputing missing values

```{r echo =TRUE}
countNA <- sum(!complete.cases(data))
```  
There is a total of **`r countNA`** rows with missing values.  

Further evaluation of the data shows there are full days with missing values for the steps variable. To verify this, a data set was created by subsetting the original dataset with rows containing missing values and counting the number of rows per date.
```{r echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
data_ver <- data[which(is.na(data$steps)),]
verif <- data_ver %>% group_by(date) %>% summarize(count = n())
print(as.data.frame(verif))
```  
There are 8 days each containing a total of 288 rows of missing values. This number corresponds to the number of intervals recorded each day. This evaluation was performed to ensure there were no days with both recorded and missing values. Based on this information, it was decided to impute missing values with the total mean number of steps taken.  

```{r echo=TRUE}
data2 <- read.csv(file, header = TRUE, stringsAsFactor = FALSE)
data2$steps <- ifelse(is.na(data2$steps), mean(data2$steps, na.rm = TRUE), data2$steps)
data2$date <- ymd(data2$date)
new_data <- aggregate(steps~date, data2, sum)
hist(new_data$steps, main = "Steps Taken Per Day", col = "orchid", xlab = "steps/day")
```

## New mean and median calculations.  

```{r echo=TRUE}
mean(new_data$steps)
median(new_data$steps)
```  

The data was normalized by imputting missing values with the total mean steps taken. The new **mean** and **median** are equal to the previously calculated mean of **10766.19**. Imputing missing values with the mean steps taken had no major effect on the data. The new calculated median increased slightly from the previous calculated median and the frequency around the mean increased as can be expected due to the addition of more values to the center of the distribution as can be seen from the histogram shown above.  

## Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE,warning=FALSE}
weekend <- c("Saturday", "Sunday")
data2$wkday <- factor((weekdays(data2$date) %in% weekend), levels=c(FALSE, TRUE), labels=c('weekday', 'weekend'))
wkdata <- aggregate(steps~interval+wkday, data2, FUN=mean)
library(lattice)
xyplot(steps~interval|wkday, wkdata, layout=c(1,2), type = "l")
dev.off()
```
   
